---
title: "Social Isolation: Coding Assessment"
subtitle: "Due Aug 14 @ 8PM"
author: "Kate Johnson"
date: today
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
engine: knitr
knitr:
  opts_chunk: 
    fig.align: 'center'
    # fig.width: 6
    # fig.height: 4
    message: FALSE
    cache: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE)
```


Displaying my machine information for reproducibility:
```{r}
sessionInfo()
```

```{r libraries, message=F, warning=F}
library(readr)
library(survival)
library(tidyr)
library(tidymodels)
library(dplyr)
```


Below are responses to a data assessment, on topics of determinants of cognitive aging focused on social isolation score, with data provided by Dr. Ruijia Chen, August 2026. 

Dataset (for reference): "This data set is from a study that measured social isolation with a six-item index adapted from Berkman and Syme. At
the baseline visit the study team released only the raw item responses, listed below, and did not distribute a derived score. The follow-up visits (`iso08 onward`) were scored by the study team using the same rules."

```{r}
dat <- read.csv("SocialIsolationDementiaExample.csv")
```


# Descriptive Statistics

Here I assess the data structure, variable types, and next steps to approach derivation of score variable. 


```{r}
# str(long.dat)
psych::describe(dat) %>% print(n = Inf)
```

Also to assess missingness:

```{r}
sapply(dat, function(x) sum(is.na(x)))
```


# Question 1 

*1.Construct the baseline social isolation index from the raw item responses, following the scoring rules above. Note any decisions you had to make along the way. Report the distribution: how many participants you could score, the frequency of each value, and the mean, SD, and median. Does the index look reasonable to you?* 

**Response**: We are given the construction of the index score for follow up but  not as baseline. Here, I will repeat the scoring rule to derive the baseline social isolation index from the data. 


The scoring rule: 

Scoring rules. The index is the sum of six items, each scored 1 (isolated) or 0 (not isolated).
Range 0 to 6, higher means more isolated.

1. Lives alone: 1 point if Q1 = Yes.

2. Infrequent contact with children: 1 point if Q3 is 1, 2, or 3, that is, less than monthly.

3. Infrequent contact with friends: 1 point if Q4 is 1, 2, or 3.

4. Infrequent contact with other relatives: 1 point if Q5 is 1, 2, or 3.

5. No religious participation: 1 point if Q6 = 1 (never).

6. No group participation: 1 point if Q7 = 1 (never).


```{r}
dat %>% count(livalone06) # 1 = yes
dat %>% count(nkids) # 98 <- NA
dat %>%count(kidcon06) # Q3
dat %>% count(frcon06) # Q4 
dat %>% count(relcon06) # Q5 
dat %>% count(relig06) # Q6 # 1 = never
dat %>% count(club06) # Q7
#>Response options for Q3, Q4, and Q5: 1 = Never | 2 = Less than once a year | 3 = A few times a year | 4 = One to three times a month | 5 = About once a week | 6 = Daily or almost daily | 9 = Don’t know or refused

# Convert 98 to NA
dat$nkids[dat$nkids == 98] <- NA
dat$kidcon06[dat$kidcon06 == 9] <- NA
dat$frcon06[dat$frcon06 == 9] <- NA
## Cautionary note: 7 was not included for religion but was found as a value, so safe side to make it NA
dat$relig06[dat$relig06 == 7] <- NA

## Iso-index questionnaires (all but question 2, but question 2 indiciates whether response to question 3,4, or 5):


# # Q1: Lives alone, use livalone06 directly, 
# 
# # Q2: Infrequent contact with kids:
# dat$iso_with_kids <- case_when(
#   dat$nkids == 0 ~ 0, # If no kids, then must be 0, do not add the next questions
#   kidcon06 %in% c(1,2,3) ~ 1, # 1 point, if 1,2 or 3 
#   TRUE ~ NA
# )
# 
# # Q3: Infrequent contact with friends
# dat$iso_no_friends <- ifelse(frcon06 %in% c(1,2,3), 1, 0)
# 
# # Q4: Infrequent contact with relatives
# dat$iso_relatives <- ifelse(relcon06 %in% c(1,2,3), 1, 0)
# 
# # Q5: no religious participatino
# dat$iso_religion <- ifelse(relig06 == 1, 1, 0)
# 
# # Q6 No group participation
# dat$iso_no_group <- ifelse(club06 == 1, 1, 0)

dat <- dat %>%
  mutate(
    iso_with_kids = case_when(
      nkids == 0 ~ 0, # If no kids, then must be 0, do not add the next questions
      kidcon06 %in% c(1,2,3) ~ 1, # 1 point, if 1,2 or 3 
      TRUE ~ NA_real_
      ),
  iso_no_friends =  ifelse(frcon06 %in% c(1,2,3), 1, 0),
  iso_relatives =  ifelse(relcon06 %in% c(1,2,3), 1, 0), 
  iso_religion = ifelse(relig06 == 1, 1, 0),
  iso_no_group = ifelse(club06 == 1, 1, 0)
  )
colnames(dat)

## Create the baseline score: Sum of 6:
dat <- dat %>%
  mutate(baseline_iso_score = rowSums(
    select(., livalone06, iso_with_kids, iso_no_friends, iso_relatives, iso_religion, iso_no_group)
  ))


```












