---
title: "Social Isolation: Coding Assessment"
subtitle: "Due Aug 14 @ 8PM"
author: "Kate Johnson"
date: today
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
engine: knitr
knitr:
  opts_chunk: 
    fig.align: 'center'
    # fig.width: 6
    # fig.height: 4
    message: FALSE
    cache: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE)
```


Displaying my machine information for reproducibility:
```{r}
sessionInfo()
```

```{r libraries}
library(readr)

```


Below are responses to a data assessment, on topics of determinants of cognitive aging focused on social isolation score, with data provided by Dr. Ruijia Chen, August 2026. 

Dataset (for reference): "This data set is from a study that measured social isolation with a six-item index adapted from Berkman and Syme. At
the baseline visit the study team released only the raw item responses, listed below, and did not distribute a derived score. The follow-up visits (`iso08 onward`) were scored by the study team using the same rules."

```{r}
dat <- read.csv("~/UCLA Things/RC_Code_Task/SocialIsolationDementiaExample.csv")
```











