---
title: "Social Isolation: Coding Assessment"
subtitle: "Due Aug 14 @ 8PM"
author: "Kate Johnson"
date: today
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
engine: knitr
knitr:
  opts_chunk: 
    fig.align: 'center'
    # fig.width: 6
    # fig.height: 4
    message: FALSE
    cache: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE)
```


Displaying my machine information for reproducibility:
```{r}
sessionInfo()
```

```{r libraries, message=F, warning=F}
library(readr)
library(survival)
library(tidyr)
library(tidymodels)
library(dplyr)
```


Below are responses to a data assessment, on topics of determinants of cognitive aging focused on social isolation score, with data provided by Dr. Ruijia Chen, August 2026. 

Dataset (for reference): "This data set is from a study that measured social isolation with a six-item index adapted from Berkman and Syme. At
the baseline visit the study team released only the raw item responses, listed below, and did not distribute a derived score. The follow-up visits (`iso08 onward`) were scored by the study team using the same rules."

```{r}
dat <- read.csv("SocialIsolationDementiaExample.csv")
```


# Descriptive Statistics

Here I assess the data structure, variable types, and next steps to approach derivation of score variable. 


```{r}
# str(long.dat)
psych::describe(dat) %>% print(n = Inf)
```

Also to assess missingness:

```{r}
sapply(dat, function(x) sum(is.na(x)))
```


# Question 1 

*1.Construct the baseline social isolation index from the raw item responses, following the scoring rules above. Note any decisions you had to make along the way. Report the distribution: how many participants you could score, the frequency of each value, and the mean, SD, and median. Does the index look reasonable to you?* 

**Response**: We are given the construction of the index score for follow up but  not as baseline. Here, I will repeat the scoring rule to derive the baseline social isolation index from the data. 


The scoring rule: 

Scoring rules. The index is the sum of six items, each scored 1 (isolated) or 0 (not isolated).
Range 0 to 6, higher means more isolated.

1. Lives alone: 1 point if Q1 = Yes.

2. Infrequent contact with children: 1 point if Q3 is 1, 2, or 3, that is, less than monthly.

3. Infrequent contact with friends: 1 point if Q4 is 1, 2, or 3.

4. Infrequent contact with other relatives: 1 point if Q5 is 1, 2, or 3.

5. No religious participation: 1 point if Q6 = 1 (never).

6. No group participation: 1 point if Q7 = 1 (never).


Note: Process took about 20 minute, large debugging came from mutate() and not rememebering that it takes only "=" for variable assignment, and not "<-". 

```{r}
dat %>% count(livalone06) # 1 = yes
dat %>% count(nkids) # 98 <- NA
dat %>%count(kidcon06) # Q3
dat %>% count(frcon06) # Q4 
dat %>% count(relcon06) # Q5 
dat %>% count(relig06) # Q6 # 1 = never
dat %>% count(club06) # Q7
#>Response options for Q3, Q4, and Q5: 1 = Never | 2 = Less than once a year | 3 = A few times a year | 4 = One to three times a month | 5 = About once a week | 6 = Daily or almost daily | 9 = Don’t know or refused

# Convert 98 to NA
dat$nkids[dat$nkids == 98] <- NA
dat$kidcon06[dat$kidcon06 == 9] <- NA
dat$frcon06[dat$frcon06 == 9] <- NA
## Cautionary note: 7 was not included for religion but was found as a value, so safe side to make it NA
dat$relig06[dat$relig06 == 7] <- NA

## assess missingness again

sapply(dat, function(x) sum(is.na(x)))
```

```{r}


## Iso-index questionnaires (all but question 2, but question 2 indiciates whether response to question 3,4, or 5):


# # Q1: Lives alone, use livalone06 directly, 
# 
# # Q2: Infrequent contact with kids:
# dat$iso_with_kids <- case_when(
#   dat$nkids == 0 ~ 0, # If no kids, then must be 0, do not add the next questions
#   kidcon06 %in% c(1,2,3) ~ 1, # 1 point, if 1,2 or 3 
#   TRUE ~ NA
# )
# 
# # Q3: Infrequent contact with friends
# dat$iso_no_friends <- ifelse(frcon06 %in% c(1,2,3), 1, 0)
# 
# # Q4: Infrequent contact with relatives
# dat$iso_relatives <- ifelse(relcon06 %in% c(1,2,3), 1, 0)
# 
# # Q5: no religious participatino
# dat$iso_religion <- ifelse(relig06 == 1, 1, 0)
# 
# # Q6 No group participation
# dat$iso_no_group <- ifelse(club06 == 1, 1, 0)

dat <- dat %>%
  mutate(
    iso_with_kids = case_when(
      nkids == 0 ~ 0, # If no kids, then must be 0, do not add the next questions
      kidcon06 %in% c(1,2,3) ~ 1 # 1 point, if 1,2 or 3 
      ),
  iso_no_friends =  ifelse(frcon06 %in% c(1,2,3), 1, 0),
  iso_relatives =  ifelse(relcon06 %in% c(1,2,3), 1, 0), 
  iso_religion = ifelse(relig06 == 1, 1, 0),
  iso_no_group = ifelse(club06 == 1, 1, 0)
  )
colnames(dat)

## Create the baseline score: Sum of 6:
dat <- dat %>%
  mutate(baseline_iso_score = rowSums(
    select(., livalone06, iso_with_kids, iso_no_friends, iso_relatives, iso_religion, iso_no_group),na.rm = FALSE
  ))

```


## Question 1, Part 2, describe baseline score frequency

**Reponse**: 1223 where able to be scored at baseline, mean was 2.97, median was 3, approx normally skewed, sd was approx 1.42. Index appears reasonable, however there is alot of missingness that could be addressed. Theory is from the case_when for the kid count, as that had 375 missing observations, rowSum may also be including NA in summary, leaving the baseline indicator as NA. 

```{r}
## count of number that were scorable:
sum(!is.na(dat$baseline_iso_score)) # 1223 of 2404

# Count of frequency of each value
dat %>% 
  count(baseline_iso_score)

# mean, median, and SD
summary(dat$baseline_iso_score) # gives the mean, SD, and median
sd(dat$baseline_iso_score, na.rm = T) # 1.42

# Summary by gender (out of curiosity):
(sex_summary <- dat %>%
  group_by(gender) %>%
  count(baseline_iso_score))

```



# Question 2: 


* Check the timing variables. Are survtime, dem_time, and the visit times consistent with one another? Show how you checked. What is the average length of follow-up?*


```{r}
# Check timing: ensure spacing from time06, baseline time06 = 0
times <- c("time06", "time08", "time10", "time12", "time14", "time16")

# searched "how to find the largest val out of multiple columns"
dat_last <- dat %>%
  mutate(
    last_visit = pmax(time06, time08, time10, time12, time14, time16, time18, na.rm = TRUE),
    consistent = last_visit <= survtime
  )
head(dat_last$last_visit) # did not work
dat_last %>%
  count(consistent)

## Find the average follow up length

mean(dat$survtime, na.rm = T)

# mean(dat$dem_time) not exactly follow up
  
```

Average Follow up length was 9.91 years. 



# Question 3:

* Describe the missing data in the cognitive scores. Reshape the data to long format and
tabulate missing cog by wave. Is the pattern monotone or intermittent? More than one process produces it, so identify them and report how many participants fall into each. A colleague wants to analyze only participants with all seven visits. How many is that, and what would you do instead?*


```{r}
# General format for reshaping data from wide to long/ from course: 
# dat.long = reshape(dat,
#               varying = list(12:18),
#               times =  c(1998, 2001, 2004, 2007, 2010, 2013, 2016),  
#               v.names = "measured_cogfxn",
#               idvar = "pid",
#               direction = "long")

# Using Tidyverse (had to google to remember pivot_longer exists!)

dat.long <- dat %>%
  select(id, starts_with("cog")) %>%
  pivot_longer(
    cols = starts_with("cog"),
    names_to = "wave",
    names_prefix =  "cog",
    values_to = "cog_score"
  ) 

dat.long$missing <- is.na(dat.long$cog_score)
table(dat.long$missing)
head(dat.long)
```

Next part: Is the pattern monotone or intermittent? 

**Response**: Thought process: If the pattern of missingness was monotone, then would cog_score and missingness be consistent across each time/ wave? And for intermittent, that would imply that the participant missed one of the visits and came back for a later one. THere may be multiple patterns driving missingness, in which either (1) the participant has not reached that timepoint (would need to check the start date or enrollment date of particpant to study to see if it differed by maybe that the time between waves), (2) the participant is lost to follow up, or the event (death or dementia, check for competiting risks) had occurred, (3) the participant missed a visit and came to another timepoint. This may be an interrupted timepoint, such as COVID (large one), or individual level due to illness or other relating factor why the particpant missed a visit, such as difficulty access to healthcare. Pattern could be MCAR, MNAR, or MAR.

*I must note: I am not quite sure I understand the difference strongly enough what monotone vs intermittent visit implies in this case to assess the count per participant without further reference.*


Next part:  A colleague wants to analyze only participants with all seven visits. How many is that, and what would you do instead?


```{r}
# Add number of visits and ensure number is exactly 7. I.e., in long data, include only participants with 7 completed visits
count_dat <- dat[!is.na(dat$cog18),]
nrow(count_dat) # 1324
```

There are 1324 participants with all 7 visits. I would remake the original long data table including a filter step. Or, from the long dataset, I would try to filter for all those with at least 7 observations. I would group by participant id, and include those whose count of rows per id is at least 7. 


I would also caution the investigator about use of complete case, as there may be underlying factors into why a participant missed a visit that may be indicative of the event. I suggest a sensitivity analysis comparing complete case vs at least 6 visits or use of all and incorporate that missing a visit as a covariate. 

# Question 4: 

*Estimate the association between baseline social isolation and incident dementia, using the index you built in Question 1. Write down the equation for your model and say what the parameter of interest means. Report an estimate with a confidence interval and justify your covariates* 

**Response**: Estimate the association between baseline isolation index score and dem_hx, incident dementia. Since incident dementia incorporates a time component, and we have a predictor of interest, a survival model would be ideal. Since the follow up time and outcomes are given as survtime, death, dem_dx, dem_time, we have the capacity to perform this analysis. 

Other possible analysis: we could also simply run a logistic regression whether the baseline measure of social isolation is associated with onset of dementia, but that would be more of an incidence measure of risk rather than a hazard risk ratio. 


To construct the survival/ KM curve first and the Cox Ph model where we are using baseline social isolation index as a predictor:

```{r}
# Univariate model
surv_model <- coxph(
  Surv(survtime, dem_dx) ~ baseline_iso_score, 
  data = dat
)
summary(surv_model)

# Multivariate model
surv_model_2 <- coxph(
  Surv(survtime, dem_dx) ~ baseline_iso_score + gender + age06 + sbp06 + bmi06 + depress06, 
  data = dat
)
summary(surv_model_2)




```




# Comments, Next Steps, Takeaways

- I would derive an age variable from the birth year, for the age at which dementia was diagnosed and at which timepoint that was. 

- derive or check for competing risk between death and incident dementia

- I would readjust the baseline isolation score, as I found alot of missingness in that derived score, which may have been due to a logic error early on. 


- I would add a forest plot to the final question, to better display and explain the coefficient from the cox ph model and how it compared to other covariates.

- I would check for confounding, double check the long data reshape

- I would check my cox ph assumptions that the hazard function is constant across time, I would plot the cox.zph() for the data is normal and independent assumption. 


